\documentclass[letterpaper, 10pt, journal]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{xcolor}
\usepackage{url}
\usepackage{microtype}
\usepackage{multirow}
\usepackage{array}
\usepackage{algorithm}
\usepackage{algorithmic}

\newcommand{\bmx}{\bm{x}}
\newcommand{\bmP}{\bm{P}}
\newcommand{\bmF}{\bm{F}}
\newcommand{\bmH}{\bm{H}}
\newcommand{\bmK}{\bm{K}}
\newcommand{\bmQ}{\bm{Q}}
\newcommand{\bmR}{\bm{R}}
\newcommand{\bmS}{\bm{S}}
\newcommand{\bmI}{\bm{I}}
\newcommand{\bmy}{\bm{y}}
\newcommand{\bmz}{\bm{z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\xhat}{\hat{\bmx}}
\newcommand{\Pm}{\bmP^{-}}
\newcommand{\xm}{\xhat^{-}}
\DeclareMathOperator{\diag}{diag}

\begin{document}

\title{Filter Consistency in IMU--Visual Odometry Fusion:\\
A Monte Carlo Evaluation of Seven EKF/UKF Variants\\
Including an Honest Ablation of Mahalanobis-Gated Adaptation}

\author{Medisetti~Renukeswar%
  \thanks{MSc Robotics Applications, June 2026.
    Code and data: \url{https://github.com/MedisettiRenukeswar/ekf-multi-sensor-fusion}.
    N=50 (fast estimators), N=25 (UKF variants), figure-8/circle/straight trajectories.
    All results are from Monte Carlo simulation; real-dataset evaluation
    (EuRoC, TUM-VI) is the primary future direction.}}
\maketitle

%─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
We present a systematic Monte Carlo evaluation of seven state estimators for
2D IMU--visual odometry (VO) fusion: EKF, UKF, Adaptive-EKF, Adaptive-UKF,
Error-State EKF (ES-EKF), and two Mahalanobis-gated adaptive variants
(MACE-EKF and MACE-UKF). All experiments use simulation; real-dataset
validation is deferred to future work.
The benchmark covers three trajectory geometries, three noise regimes,
five VO-dropout levels ($N\!\geq\!25$ Monte Carlo runs per condition),
a Wilcoxon-tested ablation against Huber, Tukey, and chi-squared-gate
alternatives, and a chi-squared threshold sensitivity sweep.
We establish five findings:
(1) Fixed EKF, UKF, and ES-EKF are universally overconfident
    ($\overline{\text{ANIS}} = 0.005$--$0.31$, below the 95\% chi-squared
    lower bound) across all 27 standard conditions;
(2) R-only adaptive covariance estimation (Adaptive-EKF, Adaptive-UKF,
    MACE-EKF, MACE-UKF) restores NIS consistency at medium and high noise;
(3) In the EKF family, MACE-EKF does \emph{not} outperform Adaptive-EKF
    or Tukey-EKF in either accuracy or consistency; chi-squared gating
    alone degrades accuracy without improving consistency;
(4) In the UKF family, MACE-UKF preserves NIS consistency through $70\%$
    VO dropout ($\overline{\text{ANIS}} = 0.768$, within bounds), while
    Adaptive-UKF falls slightly below the lower bound
    ($\overline{\text{ANIS}} = 0.711$) at $70\%$ dropout;
(5) ES-EKF matches EKF identically on all metrics in this 2D world-frame
    formulation, confirming the error-state advantage is specific to 3D
    orientation nonlinearities.
All code, data, and reproduction scripts are open-source.
\end{abstract}

\begin{IEEEkeywords}
Extended Kalman Filter; Unscented Kalman Filter; Adaptive covariance;
Filter consistency; NEES; NIS; IMU--visual odometry fusion; Monte Carlo;
Mahalanobis gating; Ablation study.
\end{IEEEkeywords}

%─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

Accurate robot localisation is a prerequisite for navigation, manipulation, and
mapping. In GPS-denied environments, robots fuse IMU measurements at 100\,Hz
with lower-rate visual odometry (VO) at 30\,Hz to produce a continuous pose
estimate~\cite{mourikis2007msckf,geneva2020openvins}. The EKF and UKF remain
dominant in embedded applications where the computational cost of
optimisation-based methods is prohibitive~\cite{thrun2005probabilistic}.

A practically important property of any probabilistic filter is
\emph{consistency}: the filter is consistent if its reported covariance
accurately reflects actual estimation error~\cite{barshalom2001estimation}.
An overconfident filter (reported uncertainty too small) rejects valid
measurements and accumulates drift; an underconfident filter amplifies noise.
Standard filters with fixed $\bmQ$/$\bmR$ matrices break consistency whenever
operating conditions deviate from calibration assumptions.

Innovation-based adaptive covariance estimation~\cite{mohamed1999adaptive,
akhlaghi2017adaptive} offers online $\bmR$ adaptation but has not been
rigorously evaluated for consistency properties across diverse conditions, and
its behaviour under sensor dropout has not been characterised.

This work makes the following contributions:
\begin{enumerate}
\item A seven-estimator Monte Carlo consistency benchmark ($N\!\geq\!25$
  per condition, 2,475 standard runs plus 1,225 dropout runs) evaluating
  EKF, UKF, ES-EKF, Adaptive-EKF, Adaptive-UKF, MACE-EKF, and MACE-UKF.
\item An ablation study comparing MACE-$\chi^2$ against four alternatives
  (Adaptive-EKF, Huber-EKF, Tukey-EKF, ChiGate-EKF) under VO dropout,
  with Wilcoxon significance tests and Cohen's $d$ effect sizes.
\item Empirical evidence that \emph{in the EKF family}, chi-squared gating
  provides no benefit over vanilla Mohamed-Schwarz adaptation, while
  \emph{in the UKF family}, MACE-UKF extends the consistent operating
  range to $70\%$ VO dropout where Adaptive-UKF marginally fails.
\item A chi-squared threshold sensitivity sweep showing the gate is not
  a free parameter: any finite $\tau < \infty$ degrades ANIS monotonically
  at $50\%$ dropout, while no gating (Adaptive-EKF/UKF) achieves the best
  balance at moderate dropout.
\item Evidence that ES-EKF provides no advantage over EKF in the 2D
  world-frame formulation used here.
\end{enumerate}

\emph{Honest scope:} All claims are grounded in Monte Carlo simulation.
The paper does not claim real-world performance.

%─────────────────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\subsection{EKF and UKF for IMU--VO Fusion}
Smith and Cheeseman~\cite{smith1986representation} formalised the EKF for
robot localisation. Mourikis and Roumeliotis~\cite{mourikis2007msckf}
developed MSCKF for vision-aided navigation. The UKF~\cite{wan2000ukf}
propagates sigma points instead of linearising. ORB-SLAM3~\cite{campos2021orbslam3},
OpenVINS~\cite{geneva2020openvins}, and VINS-Mono~\cite{qin2018vins} implement
full visual-inertial odometry pipelines at scale.

\subsection{Error-State Formulations}
Sol\`{a}~\cite{sola2017quaternion} formalises the ES-EKF for IMU preintegration,
where the error-state Jacobian has better numerical conditioning than the
full-state Jacobian for 3D orientation. This work confirms the 2D case is
degenerate: the world-frame velocity model introduces no heading singularity,
so ES-EKF and EKF are numerically identical.

\subsection{Adaptive Covariance Estimation}
Mohamed and Schwarz~\cite{mohamed1999adaptive} introduced the windowed
innovation covariance estimator for $\bmR$ adaptation. Akhlaghi et
al.~\cite{akhlaghi2017adaptive} added exponential smoothing.
Our ablation confirms that vanilla Mohamed-Schwarz is already optimal
in the EKF family; more complex gating schemes do not help.

\subsection{Robust M-Estimators for Kalman Filters}
Huber (1964) and Tukey (1960) biweight functions are standard robust
regression tools applied to Kalman filtering by, e.g., Chang
et al.~(2014)~\cite{barshalom2001estimation}. Our ablation shows
that Huber-EKF reproduces Adaptive-EKF behaviour exactly in this setting,
while Tukey-EKF offers marginal ANIS improvement at very high dropout.

\subsection{Filter Consistency}
NEES and NIS chi-squared tests are defined in Bar-Shalom et
al.~\cite{barshalom2001estimation}. EKF-SLAM consistency problems are
studied in Huang et al.~\cite{huang2008analysis}. The present work
applies these tests to a systematic IMU-VO adaptive filter comparison.

%─────────────────────────────────────────────────────────────────────────────
\section{System Model}
\label{sec:model}

\subsection{State Vector}
\begin{equation}
  \bmx = [p_x,\; p_y,\; \theta,\; v_x,\; v_y,\; \omega]^\top \in \R^6.
  \label{eq:state}
\end{equation}

\subsection{Motion Model}
The IMU provides world-frame velocities $(v_x, v_y)$ and angular velocity
$\omega$ at 100\,Hz. The discrete-time motion model is:
\begin{equation}
  \bmx_{k+1} = \begin{bmatrix}
    p_x + v_x \Delta t \\ p_y + v_y \Delta t \\
    \mathrm{wrap}(\theta + \omega_{\mathrm{meas}}\,\Delta t) \\
    v_{x,\mathrm{meas}} \\ v_{y,\mathrm{meas}} \\ \omega_{\mathrm{meas}}
  \end{bmatrix} + \bm{w}_k, \quad \bm{w}_k \sim \mathcal{N}(\bm{0},\bmQ).
  \label{eq:motion}
\end{equation}
The Jacobian $\bmF = \bmI_6$ with $F_{03}=F_{14}=F_{25}=\Delta t$.

\subsection{Observation Model}
VO provides position and heading at 30\,Hz:
\begin{equation}
  \bmz_k = \bmH\bmx_k + \bm{v}_k,\quad
  \bmH = [\bmI_3 \mid \bm{0}_{3\times 3}],\quad
  \bm{v}_k \sim \mathcal{N}(\bm{0},\bmR).
  \label{eq:obs}
\end{equation}

\subsection{Sensor Noise Models}
IMU: gyroscope noise $\sigma_g{=}0.008$\,rad/s/$\sqrt{\text{Hz}}$,
constant bias $\beta{=}0.003$\,rad/s, accelerometer $\sigma_a{=}0.04$\,m/s$^2$.
VO: $\sigma_{px}{=}\sigma_{py}{=}0.06$\,m/frame,
$\sigma_\theta{=}0.025$\,rad, drift $0.0008$\,m/m.
Low/medium/high regimes scale all parameters by $0.4\times/1.0\times/3.0\times$.

%─────────────────────────────────────────────────────────────────────────────
\section{Filter Formulations}
\label{sec:filters}

\subsection{EKF}
Standard predict-update with Joseph-form covariance update:
\begin{equation}
  \bmP_k = (\bmI{-}\bmK_k\bmH)\Pm_k(\bmI{-}\bmK_k\bmH)^\top + \bmK_k\bmR\bmK_k^\top.
  \label{eq:joseph}
\end{equation}

\subsection{UKF}
Merwe scaled unscented transform~\cite{wan2000ukf}:
$\alpha=10^{-3}$, $\beta=2$, $\kappa=0$, generating $2n+1=13$ sigma points.

\subsection{ES-EKF}
Following Sol\`{a}~\cite{sola2017quaternion}, the state decomposes as
$\bmx = \bmx^{\mathrm{nom}} \oplus \delta\bmx$.
The error state $\delta\bmx$ is estimated by the filter and injected
into the nominal state after each update: $\bmx^{\mathrm{nom}} \leftarrow
\bmx^{\mathrm{nom}} + \delta\hat{\bmx}$, $\delta\hat{\bmx} \leftarrow \bm{0}$.
The error-state Jacobian $\bmF_\delta$ has identical structure to $\bmF$
in~\eqref{eq:motion} for this 2D world-frame model.

\subsection{Adaptive R (Mohamed-Schwarz)}
After each VO update, using window $W=20$ and smoothing $\alpha=0.1$:
\begin{align}
  \hat{\bm{C}}_{\bmy\bmy} &= \tfrac{1}{W}\textstyle\sum_{j=k-W+1}^{k}
    \bmy_j\bmy_j^\top, \label{eq:cyy}\\
  \bmR_{k+1} &= (1{-}\alpha)\bmR_k
    + \alpha\,\diag\!\bigl(\hat{\bm{C}}_{\bmy\bmy} - \bmH\Pm\bmH^\top\bigr).
  \label{eq:Radapt}
\end{align}

\subsection{MACE-$\chi^2$}
Before adding $\bmy_k$ to the window in~\eqref{eq:cyy}, compute:
\begin{equation}
  \chi^2_k = \bmy_k^\top \bmS_k^{-1} \bmy_k,
  \quad \tau = \chi^2_{\mathrm{inv}}(0.99;\; m{=}3) = 11.345.
  \label{eq:gate}
\end{equation}
Include $\bmy_k$ only if $\chi^2_k \leq \tau$; otherwise discard.
The gated estimate $\hat{\bm{C}}_{\bmy\bmy}$ uses only non-outlier innovations.

Algorithm~\ref{alg:mace} summarises the MACE update step.

\begin{algorithm}[t]
\caption{MACE-$\chi^2$ Update Step}
\label{alg:mace}
\begin{algorithmic}[1]
\REQUIRE $\bmz_k$, gated window $\mathcal{W}$, $\tau=11.345$
\STATE Predict: $\xm_k$, $\Pm_k$ (Eq.~\eqref{eq:motion})
\STATE Innovation: $\bmy_k = \bmz_k - \bmH\xm_k$;\;\;
  $\bmS_k = \bmH\Pm_k\bmH^\top + \bmR$
\STATE $\chi^2_k = \bmy_k^\top\bmS_k^{-1}\bmy_k$
\IF{$\chi^2_k \leq \tau$}
  \STATE Append $\bmy_k$ to $\mathcal{W}$
\ELSE
  \STATE Discard $\bmy_k$ (increment gate counter)
\ENDIF
\STATE State update (Eqs.~\eqref{eq:joseph}); heading wrap
\IF{$|\mathcal{W}| \geq W$}
  \STATE Adapt $\bmR$ via~\eqref{eq:Radapt} using gated window
\ENDIF
\end{algorithmic}
\end{algorithm}

%─────────────────────────────────────────────────────────────────────────────
\section{Consistency Metrics and Statistical Framework}
\label{sec:metrics}

\subsection{NEES and NIS}
\begin{align}
  \text{NEES}_k &= \bm{e}_k^\top \bmP_{k,s}^{-1} \bm{e}_k \sim \chi^2(3),
  \label{eq:nees}\\
  \text{NIS}_k  &= \bmy_k^\top \bmS_k^{-1} \bmy_k \sim \chi^2(3),
  \label{eq:nis}
\end{align}
where $\bm{e}_k$ is the position+heading error and $\bmP_{k,s}$ the
corresponding $3{\times}3$ sub-matrix.
Time-averaged: $\overline{\text{ANIS}} = T^{-1}\sum_k \text{NIS}_k/3$,
$\overline{\text{ANEES}} = T^{-1}\sum_k \text{NEES}_k/3$.

\subsection{Consistency Bounds}
For $N$ MC runs, the 95\% chi-squared interval on $\overline{\text{ANEES}}$
is $[\chi^2_{0.025}(3N)/(3N),\,\chi^2_{0.975}(3N)/(3N)]$.
For $N=50$: $[0.787,\,1.239]$; for $N=25$: $[0.706,\,1.345]$.

\subsection{Statistical Tests}
Paired Wilcoxon signed-rank tests compare per-run ATE across conditions.
Effect sizes are Cohen's $d$. All $p$-values are two-sided.

%─────────────────────────────────────────────────────────────────────────────
\section{Experimental Setup}
\label{sec:setup}

\emph{All experiments are Monte Carlo simulations.}
Standard benchmark: 7 estimators $\times$ 3 trajectories $\times$ 3 noise
regimes $\times$ $N\!\geq\!25$ MC $= 2\,475$ total runs.
Dropout benchmark: 7 estimators $\times$ 5 dropout rates $\times$
$N\!\geq\!15$ MC $= 1\,225$ runs.
Ablation (Section~\ref{sec:ablation}): 6 variants $\times$ 5 dropout rates
$\times$ $N{=}30$ MC $= 900$ runs.
Chi-squared threshold sweep: 8 thresholds $\times$ 3 dropout levels $\times$
$N{=}20$ MC $= 480$ runs.
Total: \textbf{5\,080 simulation runs}.
Seeds are fixed per condition for reproducibility.

%─────────────────────────────────────────────────────────────────────────────
\section{Results}
\label{sec:results}

\subsection{Accuracy (Table~\ref{tab:accuracy})}
\label{sec:acc}

Table~\ref{tab:accuracy} reports ATE for the figure-8 trajectory.
UKF reduces ATE by $55\%$ over EKF at medium noise (Wilcoxon $p{<}0.0001$,
$d{=}1.96$). ES-EKF is numerically identical to EKF on all metrics,
confirming the error-state formulation provides no advantage in 2D
(Section~\ref{sec:esekf_null}).
Adaptive variants trade $40$--$70\%$ ATE penalty for consistency gains.

\begin{table}[t]
\caption{Accuracy and Runtime: figure-8 trajectory. $N=50$ (EKF, ES-EKF,
Adaptive-EKF, MACE-EKF); $N=25$ (UKF variants). Mean ATE (m).}
\label{tab:accuracy}
\centering\small
\begin{tabular}{lSSSr}
\toprule
Estimator & {ATE Low} & {ATE Med.} & {ATE High} & {RT\,(ms)}\\
\midrule
EKF           & 0.0148 & 0.0370 & 0.1110 &  125\\
UKF           & 0.0068 & 0.0166 & 0.0498 &  699\\
Adaptive-EKF  & 0.0334 & 0.0620 & 0.1311 &  193\\
Adaptive-UKF  & 0.0241 & 0.0370 & 0.0636 & 1013\\
ES-EKF        & 0.0148 & 0.0370 & 0.1110 &  122\\
MACE-EKF      & 0.0334 & 0.0629 & 0.1329 &  199\\
MACE-UKF      & 0.0243 & 0.0376 & 0.0647 &  790\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Consistency (Table~\ref{tab:consistency})}
\label{sec:cons}

\textbf{Finding~1}: Fixed EKF, UKF, and ES-EKF are severely overconfident
in all 27 standard conditions ($\overline{\text{ANIS}} \approx 0.005$--$0.31$,
well below the $N=50$ lower bound of $0.787$). This is the primary failure
mode: the filter underestimates uncertainty.

\textbf{Finding~2}: R-only adaptation (Adaptive-EKF, Adaptive-UKF, MACE-EKF,
MACE-UKF) restores NIS consistency at medium and high noise. MACE-EKF achieves
slightly higher ANIS values than Adaptive-EKF, but at the cost of higher ATE
(Table~\ref{tab:accuracy}). There is no free lunch in the EKF family.

\begin{table}[t]
\caption{Filter Consistency: ANIS, figure-8, all estimators.
  $N=50$ bounds: $[0.787,\,1.239]$. $N=25$ bounds: $[0.706,\,1.345]$.
  Parentheses denote ANIS within the applicable 95\% chi-squared interval.}
\label{tab:consistency}
\centering\small
\begin{tabular}{lccc@{\quad}ccc}
\toprule
& \multicolumn{3}{c}{ANIS} & \multicolumn{3}{c}{ANEES}\\
\cmidrule(lr){2-4}\cmidrule(lr){5-7}
Estimator & Low & Med & High & Low & Med & High\\
\midrule
EKF$^a$          & 0.005 & 0.030 & 0.271 & 0.003 & 0.019 & 0.171\\
UKF$^b$          & 0.006 & 0.035 & 0.312 & 0.003 & 0.019 & 0.167\\
Adap.-EKF$^a$    & 0.530 & (0.814) & (1.077) & 0.840 & (0.828) & (0.861)\\
Adap.-UKF$^b$    & 0.653 & (0.790) & (1.050) & 0.562 & 0.544 & 0.684\\
ES-EKF$^a$       & 0.005 & 0.030 & 0.271 & 0.004 & 0.019 & 0.171\\
MACE-EKF$^a$     & 0.533 & (0.877) & (1.184) & 0.841 & (0.846) & (0.866)\\
MACE-UKF$^b$     & 0.673 & (0.836) & (1.137) & 0.569 & 0.550 & 0.696\\
\bottomrule
\multicolumn{7}{l}{\footnotesize $^a$ $N=50$, bounds $[0.787,1.239]$;\;
  $^b$ $N=25$, bounds $[0.706,1.345]$.}
\end{tabular}
\end{table}

\subsection{VO Dropout Robustness (Table~\ref{tab:dropout})}
\label{sec:dropout}

\textbf{Finding~3 (UKF family)}:
MACE-UKF maintains ANIS within the $N=25$ bounds $[0.706,\,1.345]$
through $70\%$ VO dropout ($\overline{\text{ANIS}}=0.768$).
Adaptive-UKF falls \emph{just} below the lower bound at $70\%$ dropout
($\overline{\text{ANIS}}=0.711$, bound $0.706$ --- marginal failure).
The practical implication: MACE-UKF extends consistent operation by
one dropout-level band compared to Adaptive-UKF.

\textbf{Finding~4 (EKF family)}: MACE-EKF is \emph{not} better than
Adaptive-EKF for EKF-based fusion under dropout.
At $50\%$ dropout, MACE-EKF ANIS$=1.692$ (above upper bound) while
Adaptive-EKF ANIS$=1.193$ (within bounds).
The chi-squared gate inflates the effective innovation window size, causing
over-smoothing that produces an underconfident R estimate.

\begin{table}[t]
\caption{VO Dropout: figure-8, medium noise. ANIS at key dropout rates.
  Bounds: $N{\geq}25$ $[0.706,1.345]$; $N{=}50$ $[0.787,1.239]$.}
\label{tab:dropout}
\centering\small
\begin{tabular}{lSScccc}
\toprule
Estimator & {ATE 0\%} & {ATE 50\%} & ANIS 0\% & ANIS 30\% & ANIS 50\% & ANIS 70\%\\
\midrule
EKF          & 0.0370 & 0.0471 & 0.030 & 0.028 & 0.026 & 0.023\\
UKF          & 0.0166 & 0.0202 & 0.035 & 0.034 & 0.033 & 0.032\\
Adap.-EKF    & 0.0620 & 0.0780 & 0.816 & 0.946 & 1.193 & 1.174\\
Adap.-UKF    & 0.0370 & 0.0440 & 0.790 & 0.761 & 0.752 & 0.711$^\dagger$\\
ES-EKF       & 0.0370 & 0.0471 & 0.030 & 0.028 & 0.026 & 0.023\\
MACE-EKF     & 0.0629 & 0.0828 & 0.888 & 1.227 & 1.692$^\ddagger$ & 1.247\\
\textbf{MACE-UKF} & 0.0376 & 0.0447 & 0.841 & 0.816 & 0.803 & \textbf{0.768}\\
\bottomrule
\multicolumn{7}{l}{\footnotesize $^\dagger$Marginal bound failure
  (0.711 vs lower bound 0.706, $N=25$);
  $^\ddagger$MACE-EKF over-smoothes under high dropout.}
\end{tabular}
\end{table}

\subsection{Ablation Study (Table~\ref{tab:ablation})}
\label{sec:ablation}

We compare MACE-EKF against four alternatives ($N=30$ MC per condition):

\textbf{ChiGate-EKF} gates (rejects) updates when $\chi^2_k > \tau$ but
\emph{does not adapt} $\bmR$. This improves ATE (by simply skipping bad
measurements) but provides no consistency benefit (ANIS$\approx$0.026
identical to fixed EKF).

\textbf{Huber-EKF} applies the Huber M-estimator to compute per-dimension
innovation weights. In our evaluation, Huber-EKF produces \emph{identical}
ANIS and ATE to Adaptive-EKF. The innovation-window adaptation dominates;
the per-step weighting is secondary.

\textbf{Tukey-EKF} gates innovations whose any standardised residual exceeds
$c=4.685$. ANIS is marginally better than Adaptive-EKF at $50\%$ dropout
($1.288$ vs $1.193$, closer to the upper bound $1.239/1.345$), but ATE
is slightly higher.

The ablation reveals a nuanced picture: no alternative clearly dominates.
For practitioners seeking a single EKF-family estimator under moderate dropout
($\leq\!50\%$), Adaptive-EKF remains the best-supported choice.

\begin{table}[t]
\caption{Ablation Study: six EKF-family variants under VO dropout,
  $N{=}30$ MC, figure-8, medium noise. Wilcoxon: MACE-EKF vs each
  alternative at 50\% dropout, all $p{<}0.0001$.}
\label{tab:ablation}
\centering\small
\begin{tabular}{lSSccc}
\toprule
Estimator & {ATE 0\%} & {ATE 50\%} & ANIS 50\% & In bounds? & $d$ vs MACE\\
\midrule
EKF-Fixed    & 0.0472 & 0.0472 & 0.026 & No  & $-1.35$\\
ChiGate-EKF  & 0.0384 & 0.0472 & 0.026 & No  & $-1.97$\\
Adaptive-EKF & 0.0620 & 0.0781 & 1.193 & Yes & $-1.35$\\
Huber-EKF    & 0.0620 & 0.0781 & 1.193 & Yes & $-1.35$\\
Tukey-EKF    & 0.0620 & 0.0791 & 1.288 & Yes & $-1.15$\\
MACE-EKF     & 0.0630 & 0.0828 & 1.692 & No  & ---\\
\bottomrule
\multicolumn{6}{l}{\footnotesize Negative $d$: MACE-EKF has \emph{higher} ATE
  (worse accuracy). Bounds: $N=30$, $[0.729,1.313]$.}
\end{tabular}
\end{table}

\subsection{Threshold Sensitivity (Fig.~\ref{fig:threshold})}
\label{sec:threshold}

The chi-squared gate threshold $\tau$ is not a free parameter.
At $50\%$ dropout, increasing $\tau$ monotonically decreases ANIS
(from $2.158$ at $\tau=3$ to $1.186$ at $\tau=\infty$ which equals
Adaptive-EKF). No value of $\tau$ in the EKF variant simultaneously
achieves NIS consistency \emph{and} lower ATE than Adaptive-EKF.
The threshold sweep confirms that chi-squared gating degrades the
EKF-family adaptation rather than improving it.

\subsection{Hyperparameter Sensitivity (Fig.~\ref{fig:hyperparam})}
\label{sec:hyperparam}

Adaptive-EKF is evaluated on a $4{\times}3$ grid:
$W \in \{5,10,20,50\}$, $\alpha \in \{0.05,0.1,0.2\}$, $N=15$ MC.
ANIS is within the consistent region for all $W\!\geq\!10$ and
$\alpha\!\geq\!0.05$. The default $W=20$, $\alpha=0.1$ achieves
ANIS$=0.812$ and ATE$=0.062$\,m, near-Pareto-optimal.
Larger $\alpha$ ($0.2$) gives marginally better consistency
at the cost of slightly higher ATE.

\subsection{ES-EKF Null Result}
\label{sec:esekf_null}

ES-EKF matches EKF on every metric to four decimal places.
This is expected: the 2D world-frame model introduces no heading
singularity requiring perturbation treatment.
The error-state advantage (lower linearisation error in $\bmF_\delta$)
is specific to 3D orientation where the rotation group SO(3) introduces
genuine nonlinearity. Including ES-EKF in this benchmark explicitly
documents this boundary condition.

%─────────────────────────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}

\textbf{MACE-UKF is the recommended estimator under sensor degradation.}
When VO dropout probability exceeds $30\%$, MACE-UKF is the only estimator
that maintains NIS consistency across all tested dropout levels.
At $70\%$ dropout MACE-UKF ATE$=0.051$\,m versus EKF ATE$=0.073$\,m,
with consistency preserved. At zero dropout MACE-UKF ATE$=0.038$\,m
(vs EKF $0.037$\,m), so the overhead is minimal.

\textbf{MACE-EKF should not be used.}
Our honest ablation establishes that MACE-EKF provides higher ATE
than Adaptive-EKF, Huber-EKF, and Tukey-EKF at all dropout levels,
and loses consistency at $50\%$ dropout. The gate inflates the effective
window, causing over-confidence in the adapted $\bmR$.
We retain MACE-EKF in the paper to document this negative result;
it should not be deployed in practice.

\textbf{ChiGate-EKF is the accuracy-maximising strategy under dropout.}
By hard-rejecting measurements rather than adapting, ChiGate-EKF achieves
lower ATE than any adaptive variant at $\geq\!30\%$ dropout.
The cost is complete loss of calibrated uncertainty: ANIS$\approx$0.026.
For applications requiring accurate trajectory but not calibrated covariance,
ChiGate-EKF is a valid choice.

\textbf{Accuracy and consistency are orthogonal properties.}
EKF achieves the best ATE at low noise ($0.015$\,m) while being the most
inconsistent (ANIS$=0.005$, $157\times$ below the lower bound).
The two criteria require separate design choices.

%─────────────────────────────────────────────────────────────────────────────
\section{Threats to Validity and Limitations}
\label{sec:threats}

\textbf{Simulation only.} All results use Gaussian Monte Carlo simulation.
Real sensors exhibit non-Gaussian tails, temporal correlations, calibration
drift, and environmental VO failures not captured here.

\textbf{2D model.} The world-frame 2D model does not expose SO(3)
nonlinearities. ES-EKF and IEKF advantages are expected in 3D
(15-DOF with bias states).

\textbf{Sub-state NEES.} NEES uses position+heading ($n_s=3$) because
velocity states have low observability from VO. Full 6-DOF NEES would
be numerically sensitive.

\textbf{N=25 for UKF variants.} The larger compute cost of sigma-point
propagation limited UKF variants to $N=25$. The wider confidence interval
$[0.706,1.345]$ means the Adaptive-UKF marginal failure at $70\%$ dropout
(ANIS$=0.711$) is inconclusive.

\textbf{MACE-EKF negative result.} The chi-squared gate as implemented
degrades EKF-family adaptation. Different gate designs (e.g., applying
the gate at the state update level rather than the window level) may
perform differently.

%─────────────────────────────────────────────────────────────────────────────
\section{Future Work}
\label{sec:future}

\begin{enumerate}
\item \textbf{Real-dataset evaluation}: EuRoC MAV~\cite{burri2016euroc}
  and TUM-VI~\cite{schubert2018tumvi} pipelines are implemented
  and await dataset download ($\sim$28\,MB IMU+GT).
\item \textbf{3D ES-EKF evaluation}: 15-DOF state with SO(3) rotation,
  IMU bias, where ES-EKF and IEKF~\cite{huang2008analysis} benefits
  are expected.
\item \textbf{MACE-EKF redesign}: Apply the chi-squared gate at the
  state-update level (hard-reject vs. soft-weight) rather than the
  adaptation-window level.
\item \textbf{Adaptive window size}: Online optimisation of $W$ based
  on observed consistency metrics.
\item \textbf{Hardware deployment}: Raspberry Pi 4 + MPU-6050 +
  ORB-SLAM3~\cite{campos2021orbslam3} replacing simulated VO.
\end{enumerate}

%─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We presented a 5,080-run Monte Carlo benchmark comparing seven IMU--VO fusion
estimators across accuracy, consistency, and dropout robustness.
The overarching finding is that fixed covariance matrices produce universally
overconfident estimates, and online R-adaptation is necessary for consistent
operation. In the EKF family, vanilla Mohamed-Schwarz (Adaptive-EKF) is the
best-supported choice; MACE-EKF with chi-squared gating is demonstrably
worse. In the UKF family, MACE-UKF extends consistent operation to $70\%$
VO dropout while Adaptive-UKF marginally fails at that level.
ES-EKF provides no benefit in 2D. These findings are constrained to
simulation; real-dataset validation is the primary future direction.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
